### 2018-03-11 paper/sexSelectionSpacing-ver1.tex
Discuss high education duration and sex ratio results

\title{Son Preference, Birth Spacing, and Sex-Selective Abortions%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of Aarhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support from the University of Washington Royalty Research Fund and the 
Development Research Group of the World Bank is gratefully acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}
}

@@

Spacing, by itself, can therefore no longer be used as a direct measure of 
son preference.
However, understanding birth spacing remains a critical undertaking.
First, birth spacing is still useful in understanding son preference, 
if combined with the likelihood of observing a boy or a girl.
Second, if spacing affects health outcomes for mother and children, it is 
essential to understand what drives changes in spacing.
The better health outcomes for girls in the presence of sex selection may,
for example, be an unintended side-effect of the longer spacing that arises
from sex-selective abortions, rather than, as often assumed, be driven by a 
higher value placed on girls because of their lower number \citep{Lin2014,Hu2015}.
Third, the duration between births may be an important factor in parents' 
decisions, either for preference or economic reasons.
Whether duration plays a role in parents' decisions is especially important 
here, because even parents with strong son preference may reverse their 
decision to use prenatal sex determination, and carry the next pregnancy 
to term whether male or female, as the duration from the prior birth 
becomes sufficiently long.
Finally, we know less about what determines spacing behavior in 
developing countries than in developed countries.
With increasing numbers of women entering the labor force in developing
countries, understanding how timing decisions are made will be necessary
for the design of suitable policies \citep{Portner2018}.

@@

The spacing between births increases substantially over time, especially
for urban women. 
Urban women show an increase in median duration of between a quarter
and three-quarters of a year, whereas the increase for rural women is 
mostly around a quarter of a year.
The spacing between marriage and the first birth is still the
exception, and for this group of women there is a decline of
about a month for both urban and rural areas.

>>>>

\title{Birth Spacing in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of Aarhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}
}

@@

Spacing, by itself, is therefore no longer a valid direct measure of 
son preference, but understanding birth spacing remains a critical undertaking.
First, birth spacing is still useful in understanding son preference, 
if combined with the likelihood of observing a boy or a girl.
Second, if spacing affects health outcomes for mother and children, it is 
essential to understand what drives changes in spacing.
The better health outcomes for girls in the presence of sex selection may,
for example, be an unintended side-effect of the longer spacing that arises
from sex-selective abortions, rather than, as often assumed, be driven by a 
higher value placed on girls because of their lower number \citep{Lin2014,Hu2015}.
Third, the duration between births may be an important factor in parents' 
decisions, either for preference or economic reasons.
Whether duration plays a role in parents' decisions is especially important 
here, because even parents with strong son preference may reverse their 
decision to use prenatal sex determination, and carry the next pregnancy 
to term whether male or female, as the duration from the prior birth 
becomes sufficiently long.
Finally, we know less about what determines spacing behavior in 
developing countries than in developed countries.
With increasing numbers of women entering the labor force in developing
countries, understanding how timing decisions are made will be necessary
for the design of suitable policies \citep{Portner2018}.

@@

The spacing between births increases substantially over time, especially
for urban women. 
Urban women show an increase in median duration of between a quarter
and three-quarters of a year, whereas the increase for rural women is 
around a quarter of a year.
The spacing between marriage and the first birth is still the
exception, and for this group of women there is a decline of
about a month for both urban and rural areas.

In urban areas, the increase in spacing with only girls is so 
substantial that women with only girls have longer spacing to the 
next birth than any of the other sex compositions from 1985 onward.%
\footnote{
The exception is the second spell for the 1985--1994 period, where women
with one girl has 0.4 months longer spacing the women with one boy.
}
This reversal in the standard spacing pattern when there is strong
son preference is substantial, especially for the third and fourth 
spells.
For the latest period the difference between only girls than the
other sex compositions are between 3.4 and 5.4 months for the third
and fourth spell.
Furthermore, the differences are statistically significant, except
for only boys for the fourth spell because few women have a fourth
birth if they have three boys already.

Rural women also show a reversal of the normal spacing pattern,
although the changes are less pronounced than for urban women.
The third spell show increasingly longer spacing with only
girls compare to the other two sex compositions over time with
most differences statistically significant, but the differences
are smaller than for urban women.
For the second spell there is little difference in spacing with
one girl and with one boy and there are few rural women to 
based the fourth spell results on.

The predicted sex ratios at the end of the spells show that
the reversal in spacing patterns is not the result of a declining son
preference but rather corresponds to a substantially more male-biased 
sex ratio.
For urban women the predicted sex ratio with only girls 
is consistently above 60\% boys for the third and fourth spells.
For rural women the predicted sex ratios are lower than for urban
women, but stil substantial and statistically significant at close 
to 60\%.
The high predicted sex ratios are presumably the result of increased 
use of sex selection.

When trying to understand the strength of son preference it is interesting
that the sex ratio is also statistically significantly different
from the natural rate in the case where women already have one
son for the third and fourth spells.
Again, the sex ratio in the presence of one son for the third and
fourth spells are higher in urban areas than rural areas, although
the difference is less than for women with only girls.
Hence, it is possible that women are still willing to use sex selection
even after giving birth to one. 
Alternatively, this behavior may be in response to either experienced
or expected mortality of the first son born.

It also appears that there is a slightly elevated sex ratio among 
first-born for urban women. 
There is, however, no corresponding increase in spacing; rather
the median duration has fallen over time and is the lowest among
the three education groups.
Hence, unless duration between marriage and first birth has decreased 
substantially by itself, for example, because of improved health, it
is not clear how much stock to place on the increased sex ratio
among first-borns as evidence for sex selection among first-borns.

Both of the last two results are different from prior studies using NFHS data.
There are two potential explanations.
First, NFHS-4 is substantially larger than the three prior surveys.
Hence, it is possible that the effects have been there all along, but 
we simply did not have the power to detect them.
Second, it is possible that, at least, the first-born results are an
artifact of any remaining recall error not captured by the data restrictions.
Certain types of recall error would also explain why the sex ratio for
the second spell is statistically significantly \emph{below} the
natural sex ratio [TK cite my recall paper].
